---
output:
  html_document: default
  pdf_document: default
---
---CP 8853 Final Project"
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r}
#install.packages('tinytex')
#tinytex::install_tinytex()
```


```{r}
setwd("~/Documents/Georgia Tech/Fall 2023/CP 8853/Final Project")
```

Load libraries
```{r}
library(dplyr)
library(tmap)
library(ggplot2)
library(tidyverse)
library(readxl)
library(sf)
library(RColorBrewer)
library(viridis)
library(plotly)
library(geofacet)
library(formattable)
library(rmarkdown)

source("functions.R")
```


Your task is to compare 
(a) the energy sources of Georgia’s electricity generation and electricity CO2 emissions to 
(b) the electricity sources and emissions of three other states, and to 
(c) the electricity sources and emissions of the United States as a whole. One of your states must be in the southeastern US (Alabama, Florida, South Carolina, North Carolina, Virginia, and Tennessee) and the other two must be elsewhere in the United States.
(d) produce at least one main table comparing your areas.


1. Limit your analyses to coal, natural gas, and distillate fuel #2 (diesel) for fuel sources for electricity generation
MSN codes 
[character 1,2]: CL,NG, DF
[character 3,4]:EG, EI
[character 5]:B

2. Limit your analyses to the years 1990 to most recent available year in SEDS.


Load Complete SEDS and other datasets

```{r}
# Block 1

seds1 <- read_csv("Complete_SEDS.csv") %>% 
  tolow3()

msn1 <- read_excel("Codes_and_Descriptions.xlsx", sheet="MSN Descriptions", skip=9) %>% 
  tolow3()

stateinfo1 <- read_excel("stateinfo6.xls")

```


The st1 command loads a GIS shapefile of states.

```{r}
# Block 2
st1 <-  st_read("cb_2018_us_state_20m.shp") %>% 
  tolow3() %>% 
  mutate(statecode = stusps) %>% 
  filter(statecode != "AK" & 
           statecode != "HI" &
           statefp <= "56") %>% 
  mutate(stateabbr = statecode) %>% 
  left_join(stateinfo1)

```

Map states
```{r}
tmap_mode("plot")
st1$highlight <- ifelse(st1$statename %in% c("Alabama", "Georgia", "Indiana", "Illinois"), st1$statename, "other")

state_colors <- c(
  "Alabama" = "darkturquoise",
  "Georgia" = "orange",
  "Indiana" = "lightskyblue4",
  "Illinois" = "orchid2",
  "other" = "grey"
)

# Plot the map using tmap, setting colors based on the "highlight" column
studymap <- tm_shape(st1) +
  tm_fill(col = "highlight",
          palette = state_colors,
          title = "Case Study States")
studymap

tmap_save(studymap, "studymap.png")

```

Analysis of electricity major fuel trends by chosen states

```{r}
seds2 <-  seds1 %>% 
  left_join(stateinfo1, 
            by=c("statecode" = "stateabbr")) %>% 
  filter(msn %in% c("CLEIB", "NGEIB", "DFEIB", "TPOPP")) %>% 
  filter(year >= 1990) %>% 
  mutate(divname = ifelse(is.na(divname), "United States", divname))
```

We can now group_by and summarize for the states:

```{r}
sedsdiv3 <-  seds2 %>% 
  group_by(statecode, year, msn) %>% 
  summarize(divdata = sum(data, na.rm=TRUE))
```


Create a subset of data for 4 states chosen
```{r}
sedsdiv3sub <- sedsdiv3 %>% 
  filter(statecode %in% c("GA", "AL", "IL", "IN", "US") )
```



Calculate per capita fuel usage
```{r}
sedsdiv4 = sedsdiv3sub %>% 
  pivot_wider(names_from=msn,values_from=divdata) %>% 
  tolow() %>% 
  mutate(coal = cleib/tpopp,
         naturalgas = ngeib/tpopp,
         diesel = dfeib/tpopp)

sedsdiv4long <- sedsdiv4 %>% 
  pivot_longer(cols = 7:9,
               names_to = "fueltype",
               values_to = "genpercapita")
```


Create plots
```{r}
coalgen <- ggplot(data=sedsdiv4) +
  geom_line(aes(x=year, y=coal, col=statecode)) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("Coal per capita energy use")
coalgen

gasgen <- ggplot(data=sedsdiv4) +
  geom_line(aes(x=year, y=naturalgas, col=statecode)) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("Natural gas per capita energy use")

dieselgen <- ggplot(data=sedsdiv4) +
  geom_line(aes(x=year, y=diesel, col=statecode)) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("diesel per capita electricity generated")

ggsave("coalgen.png", coalgen, width = 6, height = 4, units = "in")
ggsave("gasgen.png", gasgen, width = 6, height = 4, units = "in")
ggsave("dieselgen.png", dieselgen, width = 6, height = 4, units = "in")
```

facet grid to show all at once
```{r}
combined_generation_grid <- ggplot(data = sedsdiv4long) +
  geom_line(aes(x = year, y = genpercapita, col = statecode), ylab = "Usage per Capita") +
  scale_color_brewer(palette = "Set2") +
  facet_grid(fueltype ~ statecode, scales = "free_y") +
  ggtitle("Electricity Generation by Fuel Type and State")
combined_generation_grid
```


Add in coefficients to get c02 emissions and pivot longer for joining w/ seds
```{r}
coef <- read_excel("coef-revised-23b.xls") %>%
  filter(row_number() <= 13) %>% 
  pivot_longer(cols = 3:35, names_to = "year", values_to = "coef")

coef$year <- as.numeric(coef$year)

```

add column to seds data with just fuel code
```{r}
sedsdiv3sub2 <- sedsdiv3sub %>% 
  mutate(fuelcode = substr(msn, 1, 2)) %>% 
  filter(msn !="TPOPP") 

```

use left join to add coef to seds, make a new emissions column, and emissions per capita
```{r}
sedsdiv5 <-  sedsdiv3sub2 %>% 
  left_join(coef, 
            by = c("fuelcode", "year")) %>% 
  mutate(emissions = divdata*coef) %>% 
  left_join(sedsdiv4, by = c("year", "statecode")) %>% 
  mutate(emitpercap = emissions/tpopp)
 
```

Make plots of emissions
```{r}
coalemit <- ggplot(data = subset(sedsdiv5, fuelname == "coal")) +
  geom_line(aes(x=year, y=emissions, col= statecode)) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("Coal CO2 Emissions")
coalemit

dieselemit <- ggplot(data = subset(sedsdiv5, fuelname == "distillate fuel oil no. 2")) +
  geom_line(aes(x=year, y=emissions, col= statecode)) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("diesel Fuel CO2 Emissions")


gasemit <- ggplot(data = subset(sedsdiv5, fuelname == "natural gas")) +
  geom_line(aes(x=year, y=emissions, col= statecode)) +
  scale_color_brewer(palette = "Set2") +
  ggtitle("Natural Gas CO2 Emissions")

ggsave("coalemit.png", coalemit, width = 6, height = 4, units = "in")
ggsave("dieselemit.png", dieselemit, width = 6, height = 4, units = "in")
ggsave("gasemit.png", gasemit, width = 6, height = 4, units = "in")
```


```{r}
combined_emissions_grid <- ggplot(data = sedsdiv5) +
  geom_line(aes(x = year, y = emissions, col = statecode)) +
  scale_color_brewer(palette = "Set2") +
  facet_grid(fuelname ~ statecode, scales = "free_y") +
  ggtitle("CO2 Emissions by Fuel Type and State")
ggsave("combined_emissions_grid.png", combined_emissions_grid, width = 8, height = 6, units = "in")
ggsave("combined_generation_grid.png", combined_generation_grid, width = 8, height = 6, units = "in")
```



create new combined emissions per state
```{r}
sedsdiv6 <- sedsdiv5 %>% 
  group_by(statecode, year) %>% 
  summarize(totemit = sum(emissions, na.rm = TRUE))
```


join shapefile to seds
```{r}
st2 <-  st1 %>% 
  right_join(sedsdiv6)
```



combined emissions for each source by state
```{r}
sedsall2 <-  seds1 %>% 
  left_join(stateinfo1, 
            by=c("statecode" = "stateabbr")) %>% 
  filter(msn %in% c("CLEIB", "NGEIB", "DFEIB", "TPOPP")) %>% 
  filter(year >= 1990) %>% 
  mutate(divname = ifelse(is.na(divname), "United States", divname))

sedsall3 <-  sedsall2 %>% 
  group_by(statecode, year, msn) %>% 
  summarize(divdata = sum(data, na.rm=TRUE))%>% 
  mutate(fuelcode = substr(msn, 1, 2)) %>% 
  filter(msn !="TPOPP") 

sedsall5 <-  sedsall3 %>% 
  left_join(coef, 
            by = c("fuelcode", "year")) %>% 
  mutate(emissions = divdata*coef) %>% 
  left_join(sedsdiv4, by = c("year", "statecode")) %>% 
  mutate(emitpercap = emissions/tpopp)

sedsall6 <- sedsall5 %>% 
  group_by(statecode, year) %>% 
  summarize(totemit = sum(emissions, na.rm = TRUE))

sedsall7 <- sedsall5 %>% 
  group_by(statecode, year) %>% 
  summarise(totepercap = sum(emitpercap))

sedsall3tp <-  sedsall2 %>% 
  group_by(statecode, year, msn) %>% 
  summarize(divdata = sum(data, na.rm=TRUE))%>% 
  mutate(fuelcode = substr(msn, 1, 2)) 

sedsall4 = sedsall3tp %>% 
  pivot_wider(names_from=msn,values_from=divdata) %>% 
  tolow() %>% 
  mutate(coal = cleib/tpopp,
         naturalgas = ngeib/tpopp,
         diesel = dfeib/tpopp)

sedsall4long <- sedsall4 %>% 
  pivot_longer(cols = 7:9,
               names_to = "fueltype",
               values_to = "genpercapita")
```

create a line chart cartogram for all states 
```{r}
geoface1 <- ggplot(sedsall6) +
  geom_line(aes(x = year, y = totemit)) +
  theme_bw() +
  facet_geo(~ statecode)
ggsave("geofacet1.png", geoface1, width = 12, height = 6, units = "in")

geofacet3 <- ggplot(sedsall5) +
  geom_line(aes(x = year, y = emissions, col = fuelname)) +
  theme_bw() +
  facet_geo(~ statecode) +
  theme(axis.text.x=element_blank()) +
  ggtitle("CO2 Emissions by Fuel Type, 1990 - 2021")
ggsave("geofacet3.png", geofacet3, width = 12, height = 6, units = "in")

geofacet4<- ggplot(sedsall4long) +
  geom_line(aes(x = year, y = genpercapita, col = fueltype), ylab = "Usage per Capita") +
  theme_bw() +
  facet_geo(~ statecode) +
  ggtitle("Electricity Generation by Fuel Type and State")
geofacet4


geofacet2 <- ggplot(data = sedsall5) +
  geom_line(aes(x = year, y = emissions, col = fuelname)) +
  scale_color_brewer(palette = "Set2") +
  facet_geo(~ statecode, scales = "free_y") +
  ggtitle("CO2 Emissions by Fuel Type and State")
geofacet2

```


